---
title: "Hamstring_Project"
author: "Page Freeman, Carlo Lopez, David Sobek, Kasey Choma, Quinn Murphy"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library libraries 
```{r}
library(tidyverse)
library(ggplot2)
#install.packages("ggridges")
library(ggridges)
library(ggExtra)
```

Load the dataset
```{r}
hamstring <- read_csv("hamstring_injuries.csv")
```
```{r}
#unique player ids - is there multiple dates from one athlete?

unique(hamstring$player_id)
#all are different athletes


table(hamstring$r_injured) #14 injuries for R
table(hamstring$l_injured) #10 injuries for L

```

Full dataset split between healthy or injured
```{r}

ham_injury <- hamstring[hamstring$r_injured == 1 | hamstring$l_injured == 1, ]

ham_healthy <- hamstring[hamstring$r_injured != 1 &  hamstring$l_injured != 1, ]

#making a duplicate for further needs 
ham_healthy1 <- ham_healthy

```

Percent difference for two test: bridge and nb (nord bord)
```{r}
ham_healthy$percent_diff_bridge <- abs(((ham_healthy$bridge_l_mean - ham_healthy$bridge_r_mean) / ((ham_healthy$bridge_l_mean + ham_healthy$bridge_r_mean) / 2))) * 100

ham_healthy$percent_diff_nb <- abs(((ham_healthy$nb_mean_l - ham_healthy$nb_mean_r)/ ((ham_healthy$nb_mean_l+ ham_healthy$nb_mean_r) /2 ))) *100


#duplicating code for ham_healthy1
ham_healthy1$percent_diff_bridge <- abs(((ham_healthy1$bridge_l_mean - ham_healthy1$bridge_r_mean) / ((ham_healthy1$bridge_l_mean + ham_healthy1$bridge_r_mean) / 2))) * 100

ham_healthy1$percent_diff_nb <- abs(((ham_healthy1$nb_mean_l - ham_healthy1$nb_mean_r)/ ((ham_healthy1$nb_mean_l+ ham_healthy1$nb_mean_r) /2 ))) *100

#duplicating code for ham_injury
ham_injury$percent_diff_bridge <- abs(((ham_injury$bridge_l_mean - ham_injury$bridge_r_mean) / ((ham_injury$bridge_l_mean + ham_injury$bridge_r_mean) / 2))) * 100

ham_injury$percent_diff_nb <- abs(((ham_injury$nb_mean_l - ham_injury$nb_mean_r)/ ((ham_injury$nb_mean_l+ ham_injury$nb_mean_r) /2 ))) *100
```

Write the csv to extract new code
```{r}
write_csv(ham_healthy, "healthy.csv")

write_csv(ham_injury, "hamstring_injury.csv")


```



Centroid-based clustering


Clustering set up 
```{r}
#loading the necessary libraries
library(cluster) # Load cluster
library(factoextra) # clustering algorithms & visualization
library(sparcl) # Sparse Clustering
library(teamcolors) # Load team colors
library(ggimage) # Load ggimage
library(ggdark) # Load ggdark


#need to scale the data first since each column and/or test used is based on a different range and we want the clustering model to pay equal attention to all of our variables

#taking out date columns, player id
hamstring <- scale(ham_healthy[ , c(3:6,8:16, 19, 20)])
# Add player id back to data frame 
ham_healthy <- cbind.data.frame(ham_healthy$player_id, hamstring)
#fixing name of player id column
names(ham_healthy)[1] <- "player_id"

```
Clustering set-up 
```{r}
set.seed(123) # Set seed for reproducibility
fit_1 <- kmeans(x = ham_healthy[,2:16], # Set data as explantory variables 
                centers = 3,  # Set number of clusters
                nstart = 25, # Set number of starts
                iter.max = 100 ) # Set maximum number of iterations to use
```


Extract the clusters
```{r}
# Extract clusters
clusters_1 <- fit_1$cluster
# Extract centers
centers_1 <- fit_1$centers

# Check samples per cluster
summary(as.factor(clusters_1))


```

```{r}
# Check players in cluster 1
cat("Cluster 1 players:\n")
ham_healthy$player_id[clusters_1 == 1]
# Check players in cluster 2
cat("Cluster 2 players:\n")
ham_healthy$player_id[clusters_1 == 2]
# Check players in cluster 3
cat("Cluster 3 players:\n")
ham_healthy$player_id[clusters_1 == 3]
```

```{r}
# Create vector of clusters
cluster <- c(1: 3)
# Extract centers
center_df <- data.frame(cluster, centers_1)

# Reshape the data
center_reshape <- gather(center_df, features, values, bridge_r_mean:percent_diff_nb)
# View first few rows
head(center_reshape)

# Create plot
g_heat_1 <- ggplot(data = center_reshape, # Set dataset
                   aes(x = features, y = cluster, fill = values)) + # Set aesthetics
  scale_y_continuous(breaks = seq(1, 4, by = 1)) + # Set y axis breaks
  geom_tile() + # Geom tile for heatmap
  coord_equal() +  # Make scale the same for both axis
  theme_set(theme_bw(base_size = 22) ) + # Set theme
  scale_fill_gradient2(low = "blue", # Choose low color
                       mid = "white", # Choose mid color
                       high = "red", # Choose high color
                       midpoint =0, # Choose mid point
                       space = "Lab", 
                       na.value ="grey", # Choose NA value
                       guide = "colourbar", # Set color bar
                       aesthetics = "fill") + # Select aesthetics to apply
  coord_flip() # Rotate plot to view names clearly
# Generate plot
g_heat_1
```
Trying 2 clusters 
```{r}
set.seed(123) # Set seed for reproducibility
fit_2 <- kmeans(x = ham_healthy[,2:16], # Set data as explantory variables 
                centers = 2,  # Set number of clusters
                nstart = 25, # Set number of starts
                iter.max = 100 )
```

Extract the Clusters
```{r}
# Extract clusters
clusters_2 <- fit_2$cluster
# Extract centers
centers_2 <- fit_2$centers

# Check samples per cluster
summary(as.factor(clusters_2))

```

```{r}
# Check players in cluster 1
cat("Cluster 1 players:\n")
ham_healthy$player_id[clusters_2 == 1]
# Check players in cluster 2
cat("Cluster 2 players:\n")
ham_healthy$player_id[clusters_2 == 2]
```

Clustering model
```{r}
# Create vector of clusters
cluster2 <- c(1:2)
# Extract centers
center_df2 <- data.frame(cluster2, centers_2)

# Reshape the data
center_reshape2 <- gather(center_df2, features, values, bridge_r_mean:percent_diff_nb)
# View first few rows
head(center_reshape2)

# Create plot
g_heat_2 <- ggplot(data = center_reshape2, # Set dataset
                   aes(x = features, y = cluster2, fill = values)) + # Set aesthetics
  scale_y_continuous(breaks = seq(1, 4, by = 1)) + # Set y axis breaks
  geom_tile() + # Geom tile for heatmap
  coord_equal() +  # Make scale the same for both axis
  theme_set(theme_bw(base_size = 22) ) + # Set theme
  scale_fill_gradient2(low = "blue", # Choose low color
                       mid = "white", # Choose mid color
                       high = "red", # Choose high color
                       midpoint =0, # Choose mid point
                       space = "Lab", 
                       na.value ="grey", # Choose NA value
                       guide = "colourbar", # Set color bar
                       aesthetics = "fill") + # Select aesthetics to apply
  coord_flip() # Rotate plot to view names clearly
# Generate plot
g_heat_2
```

Extracting all player ids of clustering model for analysis 
```{r}

#2 cluster model 

#group 1 player ids
players_to_filter1 <- c(281933, 757012, 946264, 782022,  299418,919089,746322,724684,350344,639629,952954,963980,139721,543545,897545,487115,325124199946,157872,505616, 409101 ,184817 ,366115, 921542 ,327196, 667161 ,114656 ,689480 ,356998, 861780 ,373898, 435548 ,333045, 874402 ,837613, 689771,150267, 529220 ,326946, 776572 ,191886, 279357 ,432005 ,421241, 492881, 225697, 869026 ,693833, 321731 ,374176 ,382010, 211607, 882595, 796088, 335070, 861600, 191758, 753490, 356881, 910231, 496489 ,388349, 591072, 376425, 524843, 901321 ,983842, 234177, 402584 ,440487 ,386923, 241930,947669, 241609 ,513427 ,269645, 533589, 511619 ,713286, 918382, 500940 ,499141, 862576, 611873, 705929, 213105, 276172)

#extracting out any player ids of data frame that are in "players_to_filter1"
group1 <- ham_healthy1[ham_healthy1$player_id %in% players_to_filter1, ]

#group 2 player ids 
players_to_filter2 <- c(416551,911260,620109,941212,251342,418244,417964,823654,879197,566578 ,294064 ,923409, 494992 ,840408 ,401073 ,829525, 329686 ,339318, 777513, 515656, 309143, 120797, 146593, 218985 ,919491 ,695422 ,315240 ,387939, 291120 ,767318, 350296, 674301 ,997746 ,290663, 609759 ,344807,762613, 552345 ,627117, 153335, 259443, 463027, 269668, 362766, 346058, 194137 ,487473, 657057, 907896 ,175754, 671706, 831708, 191362, 503918,330199, 279395 ,160954, 485329, 784423, 493145 ,917619 ,959494, 126856 ,839996, 876841 ,204439, 951108, 695485, 339517, 408012 ,562008 ,668519, 470287, 992413 ,349627, 366442, 160166, 148963, 213450 ,200119, 929409, 923220, 615822 ,907554, 868331 ,547292, 215880 ,956075 ,778894, 464256, 732130, 291348, 255168, 988272, 679190, 638511, 731674, 235874, 443187 ,581514, 421017 ,736862, 349038, 753710, 604378, 646334 ,901679, 787932, 756988, 314194, 208024 ,595472, 803675 ,191269 ,548065, 754290 ,918392, 194762, 323267, 543266, 981095, 726577 ,284343,488900 ,149830, 939562, 953953, 580426, 323207 ,354905 ,347518, 634423, 901783 ,498922, 182588, 521305 ,970334, 567350 ,411555, 475545, 163209, 282067, 590198, 951343,156952, 794756 ,964787, 802817, 931108, 586697, 729619 ,230863, 559306 ,409766, 726515, 173956 ,139152, 461258, 501802, 883256,846372 ,134006,294077 ,953338, 539387, 753433, 257587, 672408, 126614 ,722466 ,847376 ,640111, 523163, 659714 ,812930, 584142 ,863054, 433147)
  
#extracting out any player ids of data frame that are in "players_to_filter2"
group2 <- ham_healthy1[ham_healthy1$player_id %in% players_to_filter2, ]


```

Making an csv file for the groups
```{r}
write_csv(group1, "clust2_group1.csv")
write_csv(group2, "clust2_group2.csv")
```



Analysis of 2 cluster model
```{r}

#means are vastly different for bridge means 
#441-443
mean(group1$bridge_r_mean)
mean(group1$bridge_l_mean)

#353-354
mean(group2$bridge_r_mean)
mean(group2$bridge_l_mean)

#these are closer to group 2 - 368-370
mean(ham_injury$bridge_r_mean)
mean(ham_injury$bridge_l_mean)
```
```{r}
#489-492
mean(group1$nb_mean_l)
mean(group1$nb_mean_r)

#356-351
mean(group2$nb_mean_l)
mean(group2$nb_mean_r)

#Again, similar to group 2, 399-386
mean(ham_injury$nb_mean_l)
mean(ham_injury$nb_mean_r)
```

Bridge test is a HUGE indicator of other correlations (and biggest value of in the clustering)
- interesting since eccentric was what was talked about as "important" with Michael 

*RUN cluster groups again to get percent difference variables?

